Improving Software Maintainability through Automated Refactoring of Code Clones

### 2019-10-23 saner/paper.tex
CR = CloneRefactor

\begin{abstract}
Duplication in source code is often seen as one of the most harmful types of technical debt as it increases the size of the codebase and creates implicit dependencies between fragments of code. To remove such anti-patterns, the codebase should be refactored. Many tools aid in the detection process of such duplication problems but do not determine whether refactoring an identified clone would improve the maintainability.

@@

This paper is organized as follows. In Sec.~\ref{sec:background} we introduce important terminilogy in the field of clone refactoring. In Sec.~\ref{sec:relatedwork} we introduce the state-of-the-art in automated clone refactoring research. In Sec.~\ref{sec:research} we propose a definition for clones such that they can and should be automatically refactored. In Sec.~\ref{sec:clonerefactor} we introduce our clone refactoring tool. In Sec.~\ref{sec:experimentalsetup} we explain the setup of our experiments, of which the results are shown in Sec.\ref{sec:results} and discussed in \ref{sec:discussion}.
%We evaluate our approach by comparing the maintainability of before- and after snapshots of a codebase for each duplication problem that is refactored. We define which factors have the greatest influence on whether a clone improves maintainability when refactored. We find a set of thresholds by which clones can be detected that are more likely to improve maintainability when refactored. This allows for more a accurate suggestion of code clones for refactoring and can assist in the refactoring process.

@@

\section{A tool for automatic clone refactoring}\label{sec:clonerefactor}
To detect such clones that can be refactored, we surveyed a set of modern clone detection tools and techniques~\cite{svajlenko2014evaluating, sheneamer2016survey, gautam2016various, roy2009comparison}. We created a set of control projects to determine the suitability of these tools to detect refactorable clones, either through configuration or postprocessing of their output.

None of the surveyed tools~\cite{ragkhitwetsagul2019siamese, roy2008nicad, semura2017ccfindersw, sajnani2016sourcerercc, saini2018oreo, jiang2007deckard, kamalpriya2017enhancing, mazinanian2016jdeodorant} were suitable with our findings about refactorable clones, because of which we propose CloneRefactor. This tool goes through a 3-step process to automatically refactor clones as shown in Fig.~\ref{fig:clonerefactorflow}. In the following sections, we explain these steps.

\begin{figure}[H]
  \includegraphics[width=1\columnwidth]{img/flow}
  \caption{Simple flow diagram of CloneRefactor.}
  \label{fig:clonerefactorflow}
\end{figure}

@@

After all clones have been detected, CloneRefactor determines whether clone classes can be merged into gapped clones (see Sec.~\ref{sec:t3r}). The maximum size of the gap is limited by a threshold. This threshold is the percentage of (not-cloned) statements in the gap against the sum of statements in both clones surrounding it. Unlike the expression variability threshold, this threshold can exceed 100\%:
\begin{equation}\label{eq:type3r}
\text{Gap Size}=\frac{\text{Statements in gap}}{\text{Statements in clones}}*100
\end{equation}

@@

Based on these relations, CloneRefactor determines where to place the cloned code when refactored. These categories are mutually exclusive: a clone class will be flagged as the first relation in the above list that it applies to. The code of clones that have a \textit{Common Class} relation can be refactored by placing the cloned code in this same class. The code of clones with a \textit{Common Hierarchy} relation can be placed in the intersecting class in the hierarchy (the class all clone instances have in common as an ancestor). The code of clones with a \textit{Common Interface} relation can be placed in the intersecting interface, but in the process has to become part of the classes' public contract. The code of clones that are \textit{Unrelated} can be placed in a newly created place: either a utility class, a new superclass abstraction or an interface.

@@

\subsection{Refactoring}
CloneRefactor can refactor clones using the ``Extract Method'' refactoring technique. In this section, we show which clones we refactor and how we apply these transformations.

\subsubsection{Extract Method} \label{sec:refactorability}
Several influencing factors may obstruct the possibility to extract code to a new method:
\begin{itemize}
    \item \textbf{Complex Control Flow}: This clone contains \texttt{break}, \texttt{continue} or \texttt{return} statements, obstructing the possibility of method extraction.
    \item \textbf{Spans Part Of A Block}: This clone spans a part of a block statement. An example of this is shown in Fig.~\ref{fig:refactorabilitypartialblock}.
    \item \textbf{Is Not A Partial Method}: If the clone does not fall in the ``Partial method'' category of Sec.~\ref{sec:location}, the ``Extract Method'' refactoring technique cannot be applied.
    \item \textbf{Returns Multiple Values}: If a clone modifies or declares multiple pieces of data that it should return.
    \item \textbf{Top-Level Non-Statement}: If one of the top-level AST nodes of the clone is not a statement. For instance, if a (part of) an anonymous class is cloned.
    \item \textbf{Can Be Extracted}: This clone is a fragment of code that can directly be extracted to a new method. Then, based on the relation between the clone instances, further refactoring techniques can be used to refactor the extracted methods (for instance ``pull up method'' for clones in sibling classes).
\end{itemize}
Clones that do not fall in the \textit{Can Be Extracted} category may require additional transformations or other techniques to refactor. CloneRefactor only refactors the clones that \textit{Can Be Extracted}.

\subsubsection{AST Transformation}
CloneRefactor uses JavaParser~\cite{smith2017javaparser}: an AST-parsing library that allows to modify the AST and write it back to source code. To refactor clones, CloneRefactor creates a new method declaration and moves all statements from a clone instance in the clone class to the new method. This method is placed according to the relation between the clone instances (see Sec.~\ref{sec:relation}). CloneRefactor analyzes the source code of the extracted method and populates it with the following properties:
\begin{itemize}
  \item \textbf{Parameters}: For each variable used that is not accessible from the scope of the extracted method.
  \item \textbf{A return value}: If the method modifies or declares local data that is needed outside of its scope, or if the cloned fragments already returned data.
  \item \textbf{Thrown exception}: If the method throws an uncaught exception that is not a \texttt{RuntimeException}.
\end{itemize}
CloneRefactor then removes all cloned code and replaces it with a method call to the newly created method. In case of a return value, CloneRefactor either assigns the call result, declares it or returns it accordingly.

@@

\subsubsection{Impact on maintainability metrics} \label{sec:metrics}
CloneRefactor measures the impact on maintainability metrics of the refactored source code for each clone class that is refactored. These metrics are derived from Heitlager et al.~\cite{heitlager2007practical}. This paper defines a set of metrics to measure the maintainability of a system. For each of these metrics, risk profiles are proposed to determine the maintainability impact on the system of a whole.

@@

\subsection{Corpus}
We ran all our experiments using CloneRefactor on a corpus of open source Java projects. This corpus is derived from the corpus of a study that uses machine learning to determine the suitability of Java projects on GitHub for data analysis purposes~\cite{githubCorpus2013}.

CloneRefactor requires all libraries of the projects it analyses, to find the full paths of all referenced symbols in the source code (see Sec.~\ref{sec:t1r}). Because of that requirement, we decided to filter the corpus to only projects using the Maven build automation system. We created a set of scripts to prepare such a corpus with all dependencies included.

@@

\subsection{Tool Validation}
We have validated the correctness of CloneRefactor through unit tests and empirical validation. First, we created a set of 57 control projects to verify the correctness in many (edge) cases. These projects contain clones for each identified relation, location, and refactorability category to see whether they get correctly identified. Next, we run the tool over the corpus and manually verify samples of the acquired results. This way, we check both the correctness of the identified clones, their context, and their proposed refactoring.

We also test the correctness of the resulting code after refactoring. For this, we use a GitHub project named JFreeChart. JFreeChart has a high test coverage and working tests, which allows us to test the correctness of the program after running CloneRefactor.

@@

\subsection{Thresholds}
In our corpus, CloneRefactor has refactored 12.710 clone classes and measured the change in indicated metrics (see Sec.~\ref{sec:metrics}). Using the presented formulas (see Sec.~\ref{sec:metricformula}) we determine how the characteristics of the extracted method (see Sec.~\ref{sec:characteristics}) influence the maintainability of the resulting codebase after refactoring. In this section, we explore the data received by comparing the before- and after snapshots of the system for each separate refactoring. Using this data, we find supporting evidence regarding which thresholds are most likely to find clones that should be refactored to improve maintainability.

@@

Regarding the return type of refactored clones, we see in Table~\ref{tab:return} that this has no major impact on maintainability. A method call to the extracted method that is directly returned and no return type extracted methods are slightly more favorable than the others. We think the main reason that the ``Return'' category is on top is that when a variable is declared at the end of the cloned fragment, CloneRefactor directly returns its value and removes the declaration. This decreases the volume slightly.

>>>>

\begin{abstract}
Duplication in source code is often seen as one of the most harmful types of technical debt as it increases the size of the codebase and creates implicit dependencies between fragments of code. To remove such anti-patterns, the codebase should be refactored. Many tools aid in the de©tection process of such duplication problems but do not determine whether refactoring an identified clone would improve the maintainability.

@@

This paper is organized as follows. In Sec.~\ref{sec:background} we introduce important terminilogy in the field of clone refactoring. In Sec.~\ref{sec:relatedwork} we introduce the state-of-the-art in automated clone refactoring research. In Sec.~\ref{sec:research} we propose a definition for clones such that they can and should be automatically refactored. In Sec.~\ref{sec:clonerefactor} we introduce our clone refactoring tool. In Sec.~\ref{sec:experimentalsetup} we explain the setup of our experiments, of which the results are shown in Sec.~\ref{sec:results} and discussed in Sec.~\ref{sec:discussion}.
%We evaluate our approach by comparing the maintainability of before- and after snapshots of a codebase for each duplication problem that is refactored. We define which factors have the greatest influence on whether a clone improves maintainability when refactored. We find a set of thresholds by which clones can be detected that are more likely to improve maintainability when refactored. This allows for more a accurate suggestion of code clones for refactoring and can assist in the refactoring process.

@@

\section{A tool for automatic clone refactoring}\label{sec:clonerefactor}
To detect clones that can be automatically refactored, we surveyed a set of modern clone detection tools and techniques~\cite{svajlenko2014evaluating, sheneamer2016survey, gautam2016various, roy2009comparison}. We created a set of control projects to determine the suitability of these tools to detect automatically refactorable clones, either through configuration or postprocessing of their output.

None of the surveyed tools~\cite{ragkhitwetsagul2019siamese, roy2008nicad, semura2017ccfindersw, sajnani2016sourcerercc, saini2018oreo, jiang2007deckard, kamalpriya2017enhancing, mazinanian2016jdeodorant} were suitable with our findings about refactorable clones, because of which we propose CR\footnote{Name anonymized for the blind review, camera-ready paper will include a link to the GitHub repo of the tool.}. This tool goes through a 3-step process to automatically refactor clones as shown in Fig.~\ref{fig:clonerefactorflow}. In the following sections, we explain these steps.

\begin{figure}[H]
  \includegraphics[width=1\columnwidth]{img/flow}
  \caption{Simple flow diagram of CR.}
  \label{fig:clonerefactorflow}
\end{figure}

@@

After all clones have been detected, CR determines whether clone classes can be merged into gapped clones (see Sec.~\ref{sec:t3r}). The maximum size of the gap is limited by a threshold. This threshold is the percentage of (not-cloned) statements in the gap against the sum of statements in both clones surrounding it. Unlike the expression variability threshold, this threshold can exceed 100\%:
\begin{equation}\label{eq:type3r}
\text{Gap Size}=\frac{\text{Statements in gap}}{\text{Statements in clones}}*100
\end{equation}

@@

Based on these relations, CR determines where to place the cloned code when refactored. These categories are mutually exclusive: a clone class will be flagged as the first relation in the above list that it applies to. The code of clones that have a \textit{Common Class} relation can be refactored by placing the cloned code in this same class. The code of clones with a \textit{Common Hierarchy} relation can be placed in the intersecting class in the hierarchy (the class all clone instances have in common as an ancestor). The code of clones with a \textit{Common Interface} relation can be placed in the intersecting interface, but in the process has to become part of the classes' public contract. The code of clones that are \textit{Unrelated} can be placed in a newly created place: either a utility class, a new superclass abstraction or an interface.

@@

\subsection{Refactoring}
CR can refactor clones using the ``Extract Method'' refactoring technique. In this section, we show which clones we refactor and how we apply these transformations.

\subsubsection{Extract Method} \label{sec:refactorability}
Several influencing factors may obstruct the possibility to extract code to a new method:
\begin{itemize}
    \item \textbf{Complex Control Flow}: This clone contains \texttt{break}, \texttt{continue} or \texttt{return} statements, obstructing the possibility of method extraction.
    \item \textbf{Spans Part Of A Block}: This clone spans a part of a block statement. An example of this is shown in Fig.~\ref{fig:refactorabilitypartialblock}.
    \item \textbf{Is Not A Partial Method}: If the clone does not fall in the ``Partial method'' category of Sec.~\ref{sec:location}, the ``Extract Method'' refactoring technique cannot be applied.
    \item \textbf{Returns Multiple Values}: If a clone modifies or declares multiple pieces of data that it should return.
    \item \textbf{Top-Level Non-Statement}: If one of the top-level AST nodes of the clone is not a statement. For instance, if a (part of) an anonymous class is cloned.
    \item \textbf{Can Be Extracted}: This clone is a fragment of code that can directly be extracted to a new method. Then, based on the relation between the clone instances, further refactoring techniques can be used to refactor the extracted methods (for instance ``pull up method'' for clones in sibling classes).
\end{itemize}
Clones that do not fall in the \textit{Can Be Extracted} category may require additional transformations or other techniques to refactor. CR only refactors the clones that \textit{Can Be Extracted}.

\subsubsection{AST Transformation}
CR uses JavaParser~\cite{smith2017javaparser}: an AST-parsing library that allows to modify the AST and write it back to source code. To refactor clones, CR creates a new method declaration and moves all statements from a clone instance in the clone class to the new method. This method is placed according to the relation between the clone instances (see Sec.~\ref{sec:relation}). CR analyzes the source code of the extracted method and populates it with the following properties:
\begin{itemize}
  \item \textbf{Parameters}: For each variable used that is not accessible from the scope of the extracted method.
  \item \textbf{A return value}: If the method modifies or declares local data that is needed outside of its scope, or if the cloned fragments already returned data.
  \item \textbf{Thrown exception}: If the method throws an uncaught exception that is not a \texttt{RuntimeException}.
\end{itemize}
CR then removes all cloned code and replaces it with a method call to the newly created method. In case of a return value, CR either assigns the call result, declares it or returns it accordingly.

@@

\subsubsection{Impact on maintainability metrics} \label{sec:metrics}
CR measures the impact on maintainability metrics of the refactored source code for each clone class that is refactored. These metrics are derived from Heitlager et al.~\cite{heitlager2007practical}. This paper defines a set of metrics to measure the maintainability of a system. For each of these metrics, risk profiles are proposed to determine the maintainability impact on the system of a whole.

@@

\subsection{Corpus}
We ran all our experiments using CR on a corpus of open source Java projects. This corpus is derived from the corpus of a study that uses machine learning to determine the suitability of Java projects on GitHub for data analysis purposes~\cite{githubCorpus2013}.

CR requires all libraries of the projects it analyses, to find the full paths of all referenced symbols in the source code (see Sec.~\ref{sec:t1r}). Because of that requirement, we decided to filter the corpus to only projects using the Maven build automation system. We created a set of scripts to prepare such a corpus with all dependencies included.

@@

\subsection{Tool Validation}
We have validated the correctness of CR through unit tests and empirical validation. First, we created a set of 57 control projects to verify the correctness in many (edge) cases. These projects contain clones for each identified relation, location, and refactorability category to see whether they get correctly identified. Next, we run the tool over the corpus and manually verify samples of the acquired results. This way, we check both the correctness of the identified clones, their context, and their proposed refactoring.

We also test the correctness of the resulting code after refactoring. For this, we use a GitHub project named JFreeChart. JFreeChart has a high test coverage and working tests, which allows us to test the correctness of the program after running CR.

@@

\subsection{Thresholds}
In our corpus, CR has refactored 12.710 clone classes and measured the change in indicated metrics (see Sec.~\ref{sec:metrics}). Using the presented formulas (see Sec.~\ref{sec:metricformula}) we determine how the characteristics of the extracted method (see Sec.~\ref{sec:characteristics}) influence the maintainability of the resulting codebase after refactoring. In this section, we explore the data received by comparing the before- and after snapshots of the system for each separate refactoring. Using this data, we find supporting evidence regarding which thresholds are most likely to find clones that should be refactored to improve maintainability.

@@

Regarding the return type of refactored clones, we see in Table~\ref{tab:return} that this has no major impact on maintainability. A method call to the extracted method that is directly returned and no return type extracted methods are slightly more favorable than the others. We think the main reason that the ``Return'' category is on top is that when a variable is declared at the end of the cloned fragment, CR directly returns its value and removes the declaration. This decreases the volume slightly.
